Omix

### 2019-04-15 omix/omix.Rmd
update omix.Rmd

DepMapTCGAMakeWideDF <- function(melted_df, output_type, data_source, exclusions) {
  # Input is a 4-column melted dataframe of omic features
  # Rename columns: sample_ID, TCGA_classification, Gene, Value
  colnames(df) <- c("sample_ID", "TCGA_classification", "Gene", "Value")
  if(tolower(output_type) == "average" | tolower(output_type) == "avg") {
    # Average feature values per gene per TCGA classification
    df_avg <- df %>%
              group_by(TCGA_classification, Gene) %>%
              summarise(Mean_Value = mean(Value))
    # Make wide
    df_wide <- dcast(data = df_avg,
                     formula = Gene ~ TCGA_classification,
                     value.var = "Mean_Value",
                     fun.aggregate = mean)
    # Remove genes if argument exists
    if(!missing(exclusions)) { df_wide <- df_wide[df_wide$Gene != exclusions, ] }
    rownames(df_wide) <- df_wide$Gene
    df_wide$Gene <- NULL
    colnames(df_wide) <- paste(colnames(df_wide), data_source, sep = "_")
  }  
  if(tolower(output_type) == "proportion" | tolower(output_type) == "prop") {
    # Calc proportion of 1s over 1s + 0s per gene per TCGA classification
    df_prop <- df %>% 
               group_by(TCGA_classification, Gene) %>%
               summarise(Proportion = sum(Value) / n())
    # Make wide
    df_wide <- dcast(data = df_prop,
                     formula = Gene ~ TCGA_classification,
                     value.var = "Proportion",
                     fun.aggregate = mean)
    # Remove genes if argument exists
    if(!missing(exclusions)) { df_wide <- df_wide[df_wide$Gene != exclusions, ] }
    rownames(df_wide) <- df_wide$Gene
    df_wide$Gene <- NULL
    colnames(df_wide) <- paste(colnames(df_wide), data_source, sep = "_")
  }
  return(df_wide)
}

DepMapTCGACorrMatrix <- function(DepMap_wide_df, TCGA_wide_df, corr_method) {
  # Input is the output from DepMapTCGAMakeWideDF function for DepMap and TCGA
  wide <- bind_cols(DepMap_wide_df, TCGA_wide_df)
  # Correlate all against all
  corr <- wide %>% as.matrix %>% cor(method = corr_method) %>% as.data.frame %>% rownames_to_column(var = "var1") %>% gather(var2, value, -var1)
  # Filter self-comparisons and eliminate duplicate correlations
  corr <- filter(corr, !(grepl("DepMap", var1) & grepl("DepMap", var2)) & !(grepl("TCGA", var1) & grepl("TCGA", var2)) & grepl("TCGA", var2))
  # Data management
  colnames(corr) <- c("DepMap", "TCGA", "r")
  corr <- data.frame(apply(corr, 2, function(x) gsub("_TCGA|_DepMap","", x)))
  corr$r <- as.numeric(as.character(corr$r))
  # Set up coprrelation dataframe for plotting
  corr_wide <- dcast(corr, DepMap ~ TCGA, value.var = "r")
  rownames(corr_wide) <- corr_wide$DepMap
  corr_wide$DepMap <- NULL
  return(corr_wide)
}

DepMapTCGACorrHeatmap <- function(corr_wide_df, out_filepath, legend_title, color_breaks) {
  # Libraries
  require(ComplexHeatmap)
  require(circlize)
  # Define color scale
  col_fun <- colorRamp2(color_breaks, c("black", "red3", "yellow", "white"))
  # Legend properties list
  ht_legend_list <- list(legend_direction = "horizontal", title_position = "lefttop", legend_width = unit(1.5, "in"), border = "black")
  # Get column order to have symmetrically ordered axes
  ht4order <- Heatmap(corr_wide_df, cluster_rows = FALSE)
  # Make heatmap
  ht <- Heatmap(corr_wide_df, 
                name = legend_title,
                col = col_fun, 
                rect_gp = gpar(col = "black"),
                heatmap_legend_param = ht_legend_list,
                row_order = rev(column_order(ht4order)), 
                show_column_dend = FALSE, 
                cluster_rows = FALSE, 
                row_names_side = "left",
                row_title = "DepMap cancer cell lines",
                column_title = "TCGA tumor samples", 
                column_title_side = "bottom")
  # Save to file
  png(filename = out_filepath, width = 7.5, height = 8, units = "in", res = 500)
  draw(ht, heatmap_legend_side = "top")
  dev.off()
  # Return heatmap object for in-line rendering
  return(ht)
}

DepMapTCGAPropsScatterplots <- function(DepMap_melted_df, TCGA_melted_df, out_filepath) {
  # Input is the 4-column melted dataframe of omic features from DepMap and TCGA
  # Rename columns: sample_ID, TCGA_classification, Gene, Value
  colnames(DepMap_melted_df) <- c("sample_ID", "TCGA_classification", "Gene", "Value")
  colnames(TCGA_melted_df) <- c("sample_ID", "TCGA_classification", "Gene", "Value")
  # Calculate proportions
  DepMap_prop <- DepMap_melted_df %>% group_by(TCGA_classification, Gene) %>% summarise(Proportion = sum(Value) / n())
  TCGA_prop <- TCGA_melted_df %>% group_by(TCGA_classification, Gene) %>% summarise(Proportion = sum(Value) / n())
  # Merge proportions from DepMap and TCGA
  props <- merge(DepMap_prop, TCGA_prop, by = c("TCGA_classification", "Gene"), all = TRUE, suffixes = c("_DepMap","_TCGA"))
  # Plot faceted scatterplots
  props_scatters <- ggplot(data = props, aes(x = Proportion_TCGA, y = Proportion_DepMap)) +
    facet_wrap(~ TCGA_classification, nrow = 4) +
    geom_point(alpha = 0.3, color = "darkcyan") +
    scale_y_continuous(breaks = seq(0, 1, by = 0.2), labels = seq(0, 1, by = 0.2)) +
    scale_x_continuous(breaks = seq(0, 1, by = 0.2), labels = seq(0, 1, by = 0.2)) +
    coord_cartesian(ylim = c(0, 1), xlim = c(0, 1)) +
    geom_smooth(method = "lm", se = FALSE, lwd = 0.5, color = "darkgray") +
    stat_cor(method = "pearson", label.x = 0.85, label.y = 0.15, label.sep = "\n") +
    labs(x = "TCGA frequency", y = "DepMap frequency", caption = paste0(formatC(x = length(unique(props$Gene)), format = "d", big.mark = ","), " genes compared."))
  # Save to file
  ggsave(out_filepath, props_scatters, device = "png", dpi = 450, width = 12, height = 12, units = "in")
  # Return for in-line rendering
  return(props_scatters)
}
```

@@

Data wrangling and processing:
```{r ge_heatmap_data, eval=FALSE, warning=FALSE}
# DepMap
# TPM
ge_depmap <- read.delim("./../../htscreens_giant_files/omix/GE/GE_DepMap_filter.csv.gz", sep = ",", header = TRUE, check.names = FALSE)
colnames(ge_depmap)[colnames(ge_depmap) == "cancer"] <- "TCGA_classification"
# Mean-center scale
ge_depmap[, 4:ncol(ge_depmap)] <- apply(ge_depmap[, 4:ncol(ge_depmap)], MARGIN = 2, FUN = scale, center = TRUE, scale = TRUE)
# Melt
ge_depmap_melt <- melt(ge_depmap, id.vars = c("sample_ID", "TCGA_classification"), measure.vars = colnames(ge_depmap)[4:length(colnames(ge_depmap))], variable.name = "Gene", value.name = "TPM_scaled")
# Make wide
ge_depmap_wide <- DepMapTCGAMakeWideDF(df = ge_depmap_melt, output_type = "average", data_source = "DepMap")

# TCGA gene expression
ge_tcga <- read.delim("./../../htscreens_giant_files/omix/GE/GE_Xena_filter.csv.gz", sep = ",", header = TRUE, check.names = FALSE)
colnames(ge_tcga)[colnames(ge_tcga) == "cancer"] <- "TCGA_classification"
# Mean center scale
ge_tcga[, 4:ncol(ge_tcga)] <- apply(ge_tcga[, 4:ncol(ge_tcga)], MARGIN = 2, FUN = scale, center = TRUE, scale = TRUE)
# Melt
ge_tcga_melt <- melt(ge_tcga, id.vars = c("sample_ID", "TCGA_classification"), measure.vars = colnames(ge_tcga)[4:length(colnames(ge_tcga))], variable.name = "Gene", value.name = "TPM_scaled")
# Make wide
ge_tcga_wide <- DepMapTCGAMakeWideDF(df = ge_tcga_melt, output_type = "average", data_source = "TCGA")

# Correlate all against all
ge_corr_wide_pearson <- DepMapTCGACorrMatrix(DepMap_wide_df = ge_depmap_wide, TCGA_wide_df = ge_tcga_wide, corr_method = "pearson")
ge_corr_wide_spearman <- DepMapTCGACorrMatrix(DepMap_wide_df = ge_depmap_wide, TCGA_wide_df = ge_tcga_wide, corr_method = "spearman")
```

@@

Make heatmaps:
```{r ge_heatmap, eval=FALSE, fig.height=4, fig.width=3.75}
ge_ht_p <- DepMapTCGACorrHeatmap(corr_wide_df = ge_corr_wide_pearson, out_filepath = "./figures/ge_depmap_tcga_heatmap_pearson.png", legend_title = "Expression Pearson correlation", color_breaks = c(-0.5, 0, 0.5, 1))
# draw(ge_ht_p, heatmap_legend_side = "top")

ge_ht_s <- DepMapTCGACorrHeatmap(corr_wide_df = ge_corr_wide_spearman, out_filepath = "./figures/ge_depmap_tcga_heatmap_spearman.png", legend_title = "Expression Spearman correlation", color_breaks = c(-0.5, 0, 0.5, 1))
# draw(ge_ht_s, heatmap_legend_side = "top")
```

@@

Data wrangling and processing:
```{r mut_heatmap_data, eval=FALSE}
# DepMap
mut_depmap <- read.delim("./data/mutation/mutation_DepMap_filter.csv.gz", sep = ",", header = TRUE, check.names = FALSE)
mut_depmap <- unfactorize(mut_depmap)
mut_depmap[, 4:ncol(mut_depmap)] <- apply(mut_depmap[, 4:ncol(mut_depmap)], 2, function(x) ifelse(x > 1, 1, x))
colnames(mut_depmap)[colnames(mut_depmap) == "cancer"] <- "TCGA_classification"
# Melt
mut_depmap_melt <- melt(mut_depmap, id.vars = c("sample_ID", "TCGA_classification"), measure.vars = colnames(mut_depmap)[4:length(colnames(mut_depmap))], variable.name = "Gene", value.name = "Mutation_Status")
# Make wide
mut_depmap_wide <- DepMapTCGAMakeWideDF(df = mut_depmap_melt, output_type = "proportion", data_source = "DepMap", exclusions = c("TP53"))

# TCGA
mut_tcga <- read.delim("./data/mutation/mutation_Xena_filter.csv.gz", sep = ",", header = TRUE, check.names = FALSE)
mut_tcga <- unfactorize(mut_tcga)
mut_tcga[, 4:ncol(mut_tcga)] <- apply(mut_tcga[, 4:ncol(mut_tcga)], 2, function(x) ifelse(x > 1, 1, x))
colnames(mut_tcga)[colnames(mut_tcga) == "cancer"] <- "TCGA_classification"
# Melt
mut_tcga_melt <- melt(mut_tcga, id.vars = c("sample_ID", "TCGA_classification"), measure.vars = colnames(mut_tcga)[4:length(colnames(mut_tcga))], variable.name = "Gene", value.name = "Mutation_Status")
# Make wide
mut_tcga_wide <- DepMapTCGAMakeWideDF(df = mut_tcga_melt, output_type = "proportion", data_source = "TCGA", exclusions = c("TP53"))

# Correlate all against all
mut_corr_wide_pearson <- DepMapTCGACorrMatrix(DepMap_wide_df = mut_depmap_wide, TCGA_wide_df = mut_tcga_wide, corr_method = "pearson")
mut_corr_wide_spearman <- DepMapTCGACorrMatrix(DepMap_wide_df = mut_depmap_wide, TCGA_wide_df = mut_tcga_wide, corr_method = "spearman")
```

@@

For plots labeled as "5000", the 5,000 most variable genes analyzed in the gene expression heatmap were selected for the mutation analyses. Alternatively, for the plots labeled as "252", the genes in the TCGA mutation dataset were analyzed if they were mutated in >=4% annotated samples and had a mutation in at least one cancer cell line from DepMap. TP53 was removed from the heatmap, as was done in Barretina et al. (2012). For the scatterplots, TP53 was retained.

@@

Make plots:
```{r mut_plots, eval=FALSE, fig.height=4, fig.width=3.75}
mut_ht_p <- DepMapTCGACorrHeatmap(corr_wide_df = mut_corr_wide_pearson, out_filepath = "./figures/mut_depmap_tcga_heatmap_pearson_252.png", legend_title = "Mutation Pearson correlation", color_breaks = c(0, 0.3, 0.6, 0.9))
# draw(mut_ht_p, heatmap_legend_side = "top")

mut_ht_s <- DepMapTCGACorrHeatmap(corr_wide_df = mut_corr_wide_spearman, out_filepath = "./figures/mut_depmap_tcga_heatmap_spearman_252.png", legend_title = "Mutation Spearman correlation", color_breaks = c(0, 0.3, 0.5, 0.7))
# draw(mut_ht_s, heatmap_legend_side = "top")

@@

### Continuous

Continuous CN estimate data wrangling and processing:
```{r cn_heatmap_data_continuous, eval=FALSE, warning=FALSE}
# DepMap copy number
cn_depmap <- read.delim("./../../htscreens_giant_files/omix/CN/CN_cont_DepMap_filter.csv.gz", sep = ",", header = TRUE, check.names = FALSE)
colnames(cn_depmap)[colnames(cn_depmap) == "cancer"] <- "TCGA_classification"
# Mean center scale
cn_depmap[, 4:ncol(cn_depmap)] <- apply(cn_depmap[, 4:ncol(cn_depmap)], MARGIN = 2, FUN = scale, center = TRUE, scale = TRUE)
# Melt
cn_depmap_melt <- melt(cn_depmap, id.vars = c("sample_ID", "TCGA_classification"), measure.vars = colnames(cn_depmap)[4:length(colnames(cn_depmap))], variable.name = "Gene", value.name = "CN_scaled")
cn_depmap_melt <- filter(cn_depmap_melt, TCGA_classification %in% tcga_classifications_16)
# Make wide
cn_depmap_wide <- DepMapTCGAMakeWideDF(df = cn_depmap_melt, output_type = "average", data_source = "DepMap")

# TCGA cop number
cn_tcga <- read.delim("./../../htscreens_giant_files/omix/CN/CN_cont_Xena_filter.csv.gz", sep = ",", header = TRUE, check.names = FALSE)
colnames(cn_tcga)[colnames(cn_tcga) == "cancer"] <- "TCGA_classification"
# Mean center scale
cn_tcga[, 4:ncol(cn_tcga)] <- apply(cn_tcga[, 4:ncol(cn_tcga)], MARGIN = 2, FUN = scale, center = TRUE, scale = TRUE)
# Melt
cn_tcga_melt <- melt(cn_tcga, id.vars = c("sample_ID", "TCGA_classification"), measure.vars = colnames(cn_tcga)[4:length(colnames(cn_tcga))], variable.name = "Gene", value.name = "CN_scaled")
cn_tcga_melt <- filter(cn_tcga_melt, TCGA_classification %in% tcga_classifications_16)
# Make wide
cn_tcga_wide <- DepMapTCGAMakeWideDF(df = cn_tcga_melt, output_type = "proportion", data_source = "TCGA")

# Correlate all against all
cn_corr_wide_pearson <- DepMapTCGACorrMatrix(DepMap_wide_df = cn_depmap_wide, TCGA_wide_df = cn_tcga_wide, corr_method = "pearson")
cn_corr_wide_spearman <- DepMapTCGACorrMatrix(DepMap_wide_df = cn_depmap_wide, TCGA_wide_df = cn_tcga_wide, corr_method = "spearman")
```

@@

Make plots:
```{r cn_plots, eval=TRUE, fig.height=4, fig.width=3.75}
cn_ht_p <- DepMapTCGACorrHeatmap(corr_wide_df = cn_corr_wide_pearson, out_filepath = "./figures/cn_depmap_tcga_heatmap_pearson_thresholdedloss.png", legend_title = "Copy number Pearson correlation", color_breaks = c(0, 0.2, 0.4, 0.6))
# draw(cn_ht_p, heatmap_legend_side = "top")

>>>>

DepMapTCGAMakeWideDF <- function(melted_df, output_type, data_source, exclusions) {
  # Input is a 4-column melted dataframe of omic features
  # Rename columns: sample_ID, TCGA_classification, Gene, Value
  colnames(melted_df) <- c("sample_ID", "TCGA_classification", "Gene", "Value")
  if(tolower(output_type) == "average" | tolower(output_type) == "avg") {
    # Average feature values per gene per TCGA classification
    df_avg <- melted_df %>%
              group_by(TCGA_classification, Gene) %>%
              summarise(Mean_Value = mean(Value))
    # Make wide
    df_wide <- dcast(data = df_avg,
                     formula = Gene ~ TCGA_classification,
                     value.var = "Mean_Value",
                     fun.aggregate = mean)
    # Remove genes if argument exists
    if(!missing(exclusions)) { df_wide <- df_wide[df_wide$Gene != exclusions, ] }
    rownames(df_wide) <- df_wide$Gene
    df_wide$Gene <- NULL
    colnames(df_wide) <- paste(colnames(df_wide), data_source, sep = "_")
  }  
  if(tolower(output_type) == "proportion" | tolower(output_type) == "prop") {
    # Calc proportion of 1s over 1s + 0s per gene per TCGA classification
    df_prop <- melted_df %>% 
               group_by(TCGA_classification, Gene) %>%
               summarise(Proportion = sum(Value) / n())
    # Make wide
    df_wide <- dcast(data = df_prop,
                     formula = Gene ~ TCGA_classification,
                     value.var = "Proportion",
                     fun.aggregate = mean)
    # Remove genes if argument exists
    if(!missing(exclusions)) { df_wide <- df_wide[df_wide$Gene != exclusions, ] }
    rownames(df_wide) <- df_wide$Gene
    df_wide$Gene <- NULL
    colnames(df_wide) <- paste(colnames(df_wide), data_source, sep = "_")
  }
  return(df_wide)
}

DepMapTCGACorrMatrix <- function(DepMap_wide_df, TCGA_wide_df, corr_method, data_format) {
  # Input is the output from DepMapTCGAMakeWideDF function for DepMap and TCGA
  wide <- bind_cols(DepMap_wide_df, TCGA_wide_df)
  # Correlate all against all
  corr <- wide %>% as.matrix %>% cor(method = corr_method, use = "pairwise.complete.obs") %>% as.data.frame %>% rownames_to_column(var = "var1") %>% gather(var2, value, -var1)
  # Filter self-comparisons and eliminate duplicate correlations
  corr <- filter(corr, !(grepl("DepMap", var1) & grepl("DepMap", var2)) & !(grepl("TCGA", var1) & grepl("TCGA", var2)) & grepl("TCGA", var2))
  # Data management
  colnames(corr) <- c("DepMap", "TCGA", "r")
  corr <- data.frame(apply(corr, 2, function(x) gsub("_TCGA|_DepMap","", x)))
  corr$r <- as.numeric(as.character(corr$r))
  if(data_format == "long") {
    # Leave dataframe in long format (for plotting with GGPlot)
    corr_df <- corr
  }
  if(data_format == "wide") {
    # Make wide dataframe (for plotting with ComplexHeatmap)
    corr_df <- dcast(corr, DepMap ~ TCGA, value.var = "r")
    rownames(corr_df) <- corr_df$DepMap
    corr_df$DepMap <- NULL
  }
  return(corr_df)
}

DepMapTCGACorrHeatmap <- function(corr_wide_df, out_filepath, legend_title, color_breaks) {
  # Libraries
  require(ComplexHeatmap)
  require(circlize)
  # Define color scale
  col_fun <- colorRamp2(color_breaks, c("black", "red3", "yellow", "white"))
  # Legend properties list
  ht_legend_list <- list(legend_direction = "horizontal", title_position = "lefttop", legend_width = unit(1.5, "in"), border = "black")
  # Get clustered column order to have symmetrically ordered axes
  # By default, clustering is performed by hclust()
  ht4order <- Heatmap(corr_wide_df, cluster_rows = FALSE)
  # Make heatmap
  ht <- Heatmap(corr_wide_df, 
                name = legend_title,
                col = col_fun, 
                rect_gp = gpar(col = "black"),
                heatmap_legend_param = ht_legend_list,
                row_order = rev(column_order(ht4order)), 
                show_column_dend = FALSE, 
                cluster_rows = FALSE, 
                row_names_side = "left",
                row_title = "DepMap cancer cell lines",
                column_title = "TCGA tumor samples", 
                column_title_side = "bottom")
  # Save to file
  png(filename = out_filepath, width = 7.5, height = 8, units = "in", res = 500)
  draw(ht, heatmap_legend_side = "top")
  dev.off()
  # Return heatmap object for in-line rendering
  return(ht)
}

DepMapTCGAPropsScatterplots <- function(DepMap_melted_df, TCGA_melted_df, out_filepath) {
  # Input is the 4-column melted dataframe of omic features from DepMap and TCGA
  # Rename columns: sample_ID, TCGA_classification, Gene, Value
  colnames(DepMap_melted_df) <- c("sample_ID", "TCGA_classification", "Gene", "Value")
  colnames(TCGA_melted_df) <- c("sample_ID", "TCGA_classification", "Gene", "Value")
  # Calculate proportions
  DepMap_prop <- DepMap_melted_df %>% group_by(TCGA_classification, Gene) %>% summarise(Proportion = sum(Value) / n(), N = n())
  TCGA_prop <- TCGA_melted_df %>% group_by(TCGA_classification, Gene) %>% summarise(Proportion = sum(Value) / n(), N = n())
  # Merge proportions from DepMap and TCGA
  props <- merge(DepMap_prop, TCGA_prop, by = c("TCGA_classification", "Gene"), all = TRUE, suffixes = c("_DepMap","_TCGA"))
  props$N_summary <- with(props, paste0("N_DepMap = ", N_DepMap, ", N_TCGA = ", N_TCGA))
  # Plot faceted scatterplots
  props_scatters <- ggplot(data = props, aes(x = Proportion_TCGA, y = Proportion_DepMap)) +
    facet_wrap(~ TCGA_classification * N_summary, nrow = 4) +
    geom_point(alpha = 0.3, color = "darkcyan") +
    scale_y_continuous(breaks = seq(0, 1, by = 0.2), labels = seq(0, 1, by = 0.2)) +
    scale_x_continuous(breaks = seq(0, 1, by = 0.2), labels = seq(0, 1, by = 0.2)) +
    coord_cartesian(ylim = c(0, 1), xlim = c(0, 1)) +
    geom_smooth(method = "lm", se = FALSE, lwd = 0.5, color = "darkgray") +
    stat_cor(method = "pearson", label.x = 0.85, label.y = 0.15, label.sep = "\n") +
    theme(strip.text.x = element_text(size = 12), axis.title = element_text(size = 14)) +
    labs(x = "TCGA frequency", y = "DepMap frequency", caption = paste0(formatC(x = length(unique(props$Gene)), format = "d", big.mark = ","), " genes compared."))
  # Save to file
  ggsave(out_filepath, props_scatters, device = "png", dpi = 450, width = 12, height = 13, units = "in")
  # Return for in-line rendering
  return(props_scatters)
}
```

@@

Data wrangling and processing:
```{r ge_heatmap_data, eval=FALSE, warning=FALSE}
# DepMap
# TPM
ge_depmap <- read.delim("./../../htscreens_giant_files/omix/GE/GE_DepMap_filter.csv.gz", sep = ",", header = TRUE, check.names = FALSE)
colnames(ge_depmap)[colnames(ge_depmap) == "cancer"] <- "TCGA_classification"
# Mean-center scale
ge_depmap[, 4:ncol(ge_depmap)] <- apply(ge_depmap[, 4:ncol(ge_depmap)], MARGIN = 2, FUN = scale, center = TRUE, scale = TRUE)
# Melt
ge_depmap_melt <- melt(ge_depmap, id.vars = c("sample_ID", "TCGA_classification"), measure.vars = colnames(ge_depmap)[4:length(colnames(ge_depmap))], variable.name = "Gene", value.name = "TPM_scaled")
# Make wide
ge_depmap_wide <- DepMapTCGAMakeWideDF(melted_df = ge_depmap_melt, output_type = "average", data_source = "DepMap")

# TCGA gene expression
ge_tcga <- read.delim("./../../htscreens_giant_files/omix/GE/GE_Xena_filter.csv.gz", sep = ",", header = TRUE, check.names = FALSE)
colnames(ge_tcga)[colnames(ge_tcga) == "cancer"] <- "TCGA_classification"
# Mean center scale
ge_tcga[, 4:ncol(ge_tcga)] <- apply(ge_tcga[, 4:ncol(ge_tcga)], MARGIN = 2, FUN = scale, center = TRUE, scale = TRUE)
# Melt
ge_tcga_melt <- melt(ge_tcga, id.vars = c("sample_ID", "TCGA_classification"), measure.vars = colnames(ge_tcga)[4:length(colnames(ge_tcga))], variable.name = "Gene", value.name = "TPM_scaled")
# Make wide
ge_tcga_wide <- DepMapTCGAMakeWideDF(melted_df = ge_tcga_melt, output_type = "average", data_source = "TCGA")

# Correlate all against all
ge_corr_wide_pearson <- DepMapTCGACorrMatrix(DepMap_wide_df = ge_depmap_wide, TCGA_wide_df = ge_tcga_wide, corr_method = "pearson", data_format = "wide")
```

@@

Make heatmaps:
```{r ge_heatmap, eval=FALSE, fig.height=4, fig.width=3.75}
ge_ht_p <- DepMapTCGACorrHeatmap(corr_wide_df = ge_corr_wide_pearson, out_filepath = "./figures/ge_depmap_tcga_heatmap_pearson.png", legend_title = "Expression Pearson correlation", color_breaks = c(-0.5, 0, 0.5, 1))
# draw(ge_ht_p, heatmap_legend_side = "top")
```

@@

Data wrangling and processing:
```{r mut_heatmap_data, eval=FALSE}
# DepMap
mut_depmap <- read.delim("./data/mutation/mutation_DepMap_filter.csv.gz", sep = ",", header = TRUE, check.names = FALSE)
mut_depmap <- unfactorize(mut_depmap)
mut_depmap[, 4:ncol(mut_depmap)] <- apply(mut_depmap[, 4:ncol(mut_depmap)], 2, function(x) ifelse(x > 1, 1, x))
colnames(mut_depmap)[colnames(mut_depmap) == "cancer"] <- "TCGA_classification"
# Melt
mut_depmap_melt <- melt(mut_depmap, id.vars = c("sample_ID", "TCGA_classification"), measure.vars = colnames(mut_depmap)[4:length(colnames(mut_depmap))], variable.name = "Gene", value.name = "Mutation_Status")
# Make wide
mut_depmap_wide <- DepMapTCGAMakeWideDF(melted_df = mut_depmap_melt, output_type = "proportion", data_source = "DepMap", exclusions = c("TP53"))
mut_depmap_wide_wTP53 <- DepMapTCGAMakeWideDF(melted_df = mut_depmap_melt, output_type = "proportion", data_source = "DepMap")

# TCGA
mut_tcga <- read.delim("./data/mutation/mutation_Xena_filter.csv.gz", sep = ",", header = TRUE, check.names = FALSE)
mut_tcga <- unfactorize(mut_tcga)
mut_tcga[, 4:ncol(mut_tcga)] <- apply(mut_tcga[, 4:ncol(mut_tcga)], 2, function(x) ifelse(x > 1, 1, x))
colnames(mut_tcga)[colnames(mut_tcga) == "cancer"] <- "TCGA_classification"
# Melt
mut_tcga_melt <- melt(mut_tcga, id.vars = c("sample_ID", "TCGA_classification"), measure.vars = colnames(mut_tcga)[4:length(colnames(mut_tcga))], variable.name = "Gene", value.name = "Mutation_Status")
# Make wide
mut_tcga_wide <- DepMapTCGAMakeWideDF(melted_df = mut_tcga_melt, output_type = "proportion", data_source = "TCGA", exclusions = c("TP53"))
mut_tcga_wide_wTP53 <- DepMapTCGAMakeWideDF(melted_df = mut_tcga_melt, output_type = "proportion", data_source = "TCGA")

# Correlate all against all
mut_corr_wide_pearson <- DepMapTCGACorrMatrix(DepMap_wide_df = mut_depmap_wide, TCGA_wide_df = mut_tcga_wide, corr_method = "pearson", data_format = "wide")
# mut_corr_wide_pearson <- DepMapTCGACorrMatrix(DepMap_wide_df = mut_depmap_wide_wTP53, TCGA_wide_df = mut_tcga_wide_wTP53, corr_method = "pearson", data_format = "wide")
```

@@

For plots labeled as "5000", the 5,000 most variable genes analyzed in the gene expression heatmap were selected for the mutation analyses. Alternatively, for the plots labeled as "252" or "253", the genes in the TCGA mutation dataset were analyzed if they were mutated in >=4% annotated samples and had a mutation in at least one cancer cell line from DepMap. TP53 was removed from the heatmap (n = 252), as was done in Barretina et al. (2012). For the scatterplots, TP53 was retained (n = 253).

@@

Make plots:
```{r mut_plots, eval=FALSE, fig.height=4, fig.width=3.75}
mut_ht_p <- DepMapTCGACorrHeatmap(corr_wide_df = mut_corr_wide_pearson, out_filepath = "./figures/mut_depmap_tcga_heatmap_pearson_253.png", legend_title = "Mutation Pearson correlation (incl. TP53)", color_breaks = c(0.15, 0.35, 0.55, 0.85))
# draw(mut_ht_p, heatmap_legend_side = "top")

@@

### Continuous (GISTIC2 TCGA)

Continuous CN estimate data wrangling and processing:
```{r cn_heatmap_data_continuous_GISTIC2, eval=FALSE, warning=FALSE}
# DepMap copy number
cn_depmap <- read.delim("./../../htscreens_giant_files/omix/CN/CN_cont_DepMap_filter.csv.gz", sep = ",", header = TRUE, check.names = FALSE)
colnames(cn_depmap)[colnames(cn_depmap) == "cancer"] <- "TCGA_classification"
# Mean center scale
cn_depmap[, 4:ncol(cn_depmap)] <- apply(cn_depmap[, 4:ncol(cn_depmap)], MARGIN = 2, FUN = scale, center = TRUE, scale = TRUE)
# Melt
cn_depmap_melt <- melt(cn_depmap, id.vars = c("sample_ID", "TCGA_classification"), measure.vars = colnames(cn_depmap)[4:length(colnames(cn_depmap))], variable.name = "Gene", value.name = "CN_scaled")
cn_depmap_melt <- filter(cn_depmap_melt, TCGA_classification %in% tcga_classifications_16)
# Make wide
cn_depmap_wide <- DepMapTCGAMakeWideDF(melted_df = cn_depmap_melt, output_type = "average", data_source = "DepMap")

# TCGA copy number
# GISTIC2 estimate
# https://xenabrowser.net/datapages/?dataset=TCGA.PANCAN.sampleMap%2FGistic2_CopyNumber_Gistic2_all_data_by_genes&host=https%3A%2F%2Ftcga.xenahubs.net&removeHub=https%3A%2F%2Fxena.treehouse.gi.ucsc.edu%3A443
cn_tcga <- read.delim("./../../htscreens_giant_files/omix/CN/CN_cont_Xena_filter.csv.gz", sep = ",", header = TRUE, check.names = FALSE)
colnames(cn_tcga)[colnames(cn_tcga) == "cancer"] <- "TCGA_classification"
# Mean center scale
cn_tcga[, 4:ncol(cn_tcga)] <- apply(cn_tcga[, 4:ncol(cn_tcga)], MARGIN = 2, FUN = scale, center = TRUE, scale = TRUE)
# Melt
cn_tcga_melt <- melt(cn_tcga, id.vars = c("sample_ID", "TCGA_classification"), measure.vars = colnames(cn_tcga)[4:length(colnames(cn_tcga))], variable.name = "Gene", value.name = "CN_scaled")
cn_tcga_melt <- filter(cn_tcga_melt, TCGA_classification %in% tcga_classifications_16)
# Make wide
cn_tcga_wide <- DepMapTCGAMakeWideDF(melted_df = cn_tcga_melt, output_type = "proportion", data_source = "TCGA")

# Correlate all against all
cn_corr_wide_pearson <- DepMapTCGACorrMatrix(DepMap_wide_df = cn_depmap_wide, TCGA_wide_df = cn_tcga_wide, corr_method = "pearson", data_format = "wide")
```

### Continuous (log(tumor/normal) TCGA)

Continuous CN estimate data wrangling and processing:
```{r cn_heatmap_data_continuous, eval=FALSE, warning=FALSE}
# DepMap copy number
cn_depmap <- read.delim("./../../htscreens_giant_files/omix/CN/CN_germ_del_DepMap_filter.csv.gz", sep = ",", header = TRUE, check.names = FALSE)
colnames(cn_depmap)[colnames(cn_depmap) == "cancer"] <- "TCGA_classification"
# Mean center scale
cn_depmap[, 4:ncol(cn_depmap)] <- apply(cn_depmap[, 4:ncol(cn_depmap)], MARGIN = 2, FUN = scale, center = TRUE, scale = TRUE)
# Melt
cn_depmap_melt <- melt(cn_depmap, id.vars = c("sample_ID", "TCGA_classification"), measure.vars = colnames(cn_depmap)[4:length(colnames(cn_depmap))], variable.name = "Gene", value.name = "CN_scaled")
cn_depmap_melt <- filter(cn_depmap_melt, TCGA_classification %in% tcga_classifications_16)
# Make wide
cn_depmap_wide <- DepMapTCGAMakeWideDF(melted_df = cn_depmap_melt, output_type = "average", data_source = "DepMap")

# TCGA copy number
# log(tumor/normal) estimate
# https://xenabrowser.net/datapages/?dataset=broad.mit.edu_PANCAN_Genome_Wide_SNP_6_whitelisted.gene.xena&host=https%3A%2F%2Fpancanatlas.xenahubs.net&removeHub=https%3A%2F%2Fxena.treehouse.gi.ucsc.edu%3A443
cn_tcga <- read.delim("./../../htscreens_giant_files/omix/CN/CN_germ_del_Xena_filter.csv.gz", sep = ",", header = TRUE, check.names = FALSE)
colnames(cn_tcga)[colnames(cn_tcga) == "cancer"] <- "TCGA_classification"
# Mean center scale
cn_tcga[, 4:ncol(cn_tcga)] <- apply(cn_tcga[, 4:ncol(cn_tcga)], MARGIN = 2, FUN = scale, center = TRUE, scale = TRUE)
# Melt
cn_tcga_melt <- melt(cn_tcga, id.vars = c("sample_ID", "TCGA_classification"), measure.vars = colnames(cn_tcga)[4:length(colnames(cn_tcga))], variable.name = "Gene", value.name = "CN_scaled")
cn_tcga_melt <- filter(cn_tcga_melt, TCGA_classification %in% tcga_classifications_16)
# Make wide
cn_tcga_wide <- DepMapTCGAMakeWideDF(melted_df = cn_tcga_melt, output_type = "proportion", data_source = "TCGA")

# Correlate all against all
cn_corr_wide_pearson <- DepMapTCGACorrMatrix(DepMap_wide_df = cn_depmap_wide, TCGA_wide_df = cn_tcga_wide, corr_method = "pearson", data_format = "wide")
```

@@

Make plots:
```{r cn_plots, eval=FALSE, fig.height=4, fig.width=3.75}
cn_ht_p <- DepMapTCGACorrHeatmap(corr_wide_df = cn_corr_wide_pearson, out_filepath = "./figures/cn_depmap_tcga_heatmap_pearson_continuous.png", legend_title = "Copy number Pearson correlation", color_breaks = c(-0.5, 0, 0.5, 1.0))
# draw(cn_ht_p, heatmap_legend_side = "top")
